---
title: "Preprocessing and quality control on expression data"
---

```{r echo=FALSE, include=FALSE}
library(SingleCellExperiment)
library(scater)
library(scran)
library(tibble)

matrix.please <- function(x) {
    m<-as.matrix(x[,-1])
    rownames(m)<-x[[1]]
    m
}

theme_personalised <- function() {
  theme(
    legend.position = "none",
    axis.title.x = element_blank(),
    axis.title.y = element_text(size=rel(1.0)),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
}
```

# Define settings

Define I/O
```{r define_io, echo=FALSE, include=FALSE}
source(here::here("settings.R"))

io$outfile <- io$sce
```

Define Options
```{r define_opts, echo=FALSE, include=FALSE}
# Minimum library size (coverage)
opts$coverage_threshold <- 1e5

# Minimum number of expressed features
opts$features_threshold <- 1000

# Maximum fraction of reads accounting for the top 50 features
opts$top50_threshold <- 0.50

# Maximum fraction of reads mapping to mithocondrial genes
opts$MT_threshold <- 0.25
```

# Load data

Load count matrix
```{r load_data, echo=FALSE}
counts <- fread(io$rna.counts) %>% matrix.please %>% .[,sample_metadata$id_rna]
```

Sanity checks
```{r}
all(sample_metadata$id_rna%in%colnames(counts))
all(colnames(counts)%in%sample_metadata$id_rna)
```

Load gene metadata
```{r load_gene_metadata, echo=FALSE}
gene_metadata <- fread(io$gene_metadata) %>%
  .[!duplicated(symbol),] %>% 
  as.tibble() %>% remove_rownames() %>% column_to_rownames("ens_id")
```

Match SCE features with the gene metadata
```{r parse_data}
genes <- rownames(gene_metadata[rownames(gene_metadata) %in% rownames(counts),])
gene_metadata <- gene_metadata[genes,]
counts <- counts[rownames(gene_metadata),]
```

Define mithocondrial genes
```{r}
mt <- rownames(gene_metadata)[gene_metadata$chr == "chrMT"]
mt <- mt[mt%in%rownames(counts)]
```

# Create SCE object

```{r echo=FALSE}
# Create rowData
rowData <- gene_metadata %>% GRanges()

# Create colData
colData <- sample_metadata %>% as.data.frame %>% tibble::remove_rownames() %>%
  tibble::column_to_rownames("id_rna") %>% .[colnames(counts),]

# create SCE object
sce <- SingleCellExperiment(assays = list(counts = as.matrix(counts)), rowData=rowData, colData=colData)
sce
```

```{r}
# sce$gene_strand <- sce$strand
# sce$strand <- NULL
```

# Plot QC metrics

```{r}
qc.metrics <- perCellQCMetrics(sce)
```

Library size
```{r}
libsize.drop <- qc.metrics$sum < opts$coverage_threshold

to.plot <- data.table(
  sample = colnames(sce), 
  size = qc.metrics$sum, 
  color = c("black","red")[as.numeric(libsize.drop)+1]
) %>% setkey(size) %>% .[,col:=size] %>% .[,sample:=factor(sample,levels=sample)]

p1 <- ggplot(to.plot, aes(x=sample, y=size, fill=color)) +
  geom_bar(stat='identity', position="dodge") +
  geom_hline(yintercept=opts$coverage_threshold, linetype="dashed") +
  labs(y="Library size") +
  theme_classic() + theme_personalised()

print(p1)
```

Number of expressed genes
```{r}
feature.drop <- qc.metrics$detected < opts$features_threshold

to.plot <- data.table(
  sample = colnames(sce), 
  features = qc.metrics$detected, 
  color = c("black","red")[as.numeric(feature.drop)+1]
  ) %>% setkey(features) %>% .[,col:=features] %>% .[,sample:=factor(sample,levels=sample)]

p2 <- ggplot(to.plot, aes(x=sample, y=features, fill=color)) +
  geom_bar(stat='identity', position="dodge") +
  geom_hline(yintercept=opts$features_threshold, linetype="dashed") +
  labs(y="Total number of expressed genes") +
  theme_classic() + theme_personalised()
print(p2)
```

Proportion of reads accounting for the top 50 features
```{r}
top50.drop <- qc.metrics$percent_top_50 > opts$top50_threshold*100

to.plot <- data.table(
  sample = colnames(sce), 
  pct = qc.metrics$percent_top_50,
  color = c("black","red")[as.numeric(top50.drop)+1]
  ) %>% setkey(pct) %>% .[,sample:=factor(sample,levels=sample)]

p3 <- ggplot(to.plot, aes(x=sample, y=pct, fill=color)) +
  geom_bar(stat='identity', position="dodge") +
  labs(x="", y="Fraction of reads accounting for the top 50 features") +
  geom_hline(yintercept=opts$top50_threshold*100, linetype="dashed") +
  theme_classic() + theme_personalised()
print(p3)
```

Mithocondrial reads
```{r}
mt.fraction <- colSums(counts(sce[mt,])) / qc.metrics$sum
mt.drop <- mt.fraction > opts$MT_threshold

to.plot <- data.table(
  sample = colnames(sce), 
  mt = mt.fraction,
  color = c("black","red")[as.numeric(mt.fraction)+1]
  ) %>% setkey(mt) %>% .[,sample:=factor(sample,levels=sample)]

p4 <- ggplot(to.plot, aes(x=sample, y=mt, fill=color)) +
  geom_bar(stat='identity', position="dodge") +
  labs(x="", y="Mithocondrial fraction") +
  geom_hline(yintercept=opts$MT_threshold, linetype="dashed") +
  theme_classic() + theme_personalised()
print(p4)
```

# Filter cells

Remove cells that do not pass QC from the SingleCellExperiment
```{r}
fail.qc <- colnames(sce)[( libsize.drop | feature.drop | top50.drop | mt.drop )]
sce <- sce[,!colnames(sce) %in% fail.qc]
```

Update sample metadata
```{r, echo=FALSE, include=TRUE}
sample_metadata[,pass_rnaQC:=ifelse(id_rna%in%fail.qc,FALSE,TRUE)]
table(sample_metadata$pass_rnaQC)

# Save updated metadata
fwrite(sample_metadata, io$metadata, sep="\t", na="NA", quote=F)
```

# Normalisation and log transformation

Scran normalisation
```{r normalisation, echo=FALSE, warnings=FALSE, include=TRUE}
sce_filt <- sce

# Temporarily remove the lowly expressed genes
sce_filt <- sce_filt[rowSums(counts(sce))>10,]

# Compute size factors without the lowly expressed genes
sf <- computeSumFactors(sce_filt, positive=TRUE) %>% sizeFactors()#, sf.out=T)

# qplot(sf, sce_filt$total_counts, log="xy", ylab="Library size (mapped reads)", xlab="Size factor")
ggplot(data.frame(sf=sf, counts=colSums(counts(sce)))) +
  geom_point(aes(x=sf,y=counts)) +
  labs(y="Library size (log)", x="Size factor (log)") +
  theme_bw() +
  theme(
    axis.title = element_text(colour="black", size=15),
    axis.text = element_text(colour="black", size=12)
  )

# Normalise and log transform with the lowly expressed genes
sizeFactors(sce) <- sf; sce$sizeFactor <- sf
sce <- logNormCounts(sce)
```

Mean vs variance plot
```{r echo=TRUE, include=TRUE}
to.plot <- data.frame(sd=apply(exprs(sce),1,sd), mean=apply(exprs(sce),1,mean))

ggplot(to.plot, aes(x=mean, y=sd)) +
  geom_point() + stat_smooth() +
  scale_color_manual(values=c("black","red")) +
  xlab('Mean') + ylab('Standard deviation') +
  theme_bw()
```
# update metadata
```{r}
sample_metadata[id_rna %in% colnames(sce), pass_rnaQC := TRUE]
sample_metadata[!(id_rna %in% colnames(sce)), pass_rnaQC := FALSE]
fwrite(sample_metadata, io$metadata, sep = "\t", na = "NA", quote = FALSE)
```


# Save

```{r save, echo=FALSE, include=FALSE}
saveRDS(sce, io$outfile)
```

