---
title: ""
---

# Define settings

Load default settings
```{r echo=TRUE, include=FALSE}
source(here::here("settings.R"))
```

I/O
```{r}
io$outdir <- paste0(io$basedir,"/metacc/exploration/multiome_peaks")
```

Options
```{r}
opts$stage_lineage <- c(
  "E3.5_ICM",
  "E4.5_Epiblast",
  "E4.5_Primitive_endoderm",
  "E5.5_Epiblast",
  "E5.5_Visceral_endoderm",
  "E6.5_Epiblast",
  "E6.5_Visceral_endoderm",
  "E6.5_Primitive_Streak",
  # "E6.5_ExE_ectoderm",
  "E7.5_Epiblast",
  "E7.5_Primitive_Streak",
  "E7.5_Ectoderm",
  "E7.5_Mesoderm",
  "E7.5_Endoderm"
)
# Define genomic contexts
opts$annos <- c(
  "multiome_peaks"
)
# opts$colors <- c(
#   "E4.5_Epiblast"="#C1CDCD",
#   "E5.5_Epiblast"="#C1CDCD",
#   "E6.5_Epiblast"="#C1CDCD",
#   "E7.5_Epiblast"="#C1CDCD",
#   "E7.5_Ectoderm"="steelblue",
#   "E6.5_Primitive_Streak"="sandybrown",
#   "E7.5_Primitive_Streak"="sandybrown",
#   "E7.5_Endoderm"="#43CD80",
#   "E7.5_Mesoderm"="#CD3278",
#   "E6.5_ExE_ectoderm"="black"
# )
```

# Load sample metadata

```{r}
# sample_metadata <- fread(io$metadata) %>% 
#   .[,stage_lineage:=paste(stage,lineage10x_2,sep="_")] %>%
#   .[stage_lineage%in%opts$stage_lineage] %>%
#   .[pass_metQC==TRUE | pass_accQC==TRUE] %>%
#   droplevels
# sample_metadata %>%
#   .[,c("stage","stage_lineage"):=list(factor(stage,levels=opts$stages), factor(stage_lineage,levels=opts$stage_lineage))]
# table(sample_metadata$lineage10x)


sample_metadata <- fread(io$metadata)%>%
   .[pass_metQC==TRUE | pass_accQC==TRUE] %>% 
  .[!is.na(celltype.mapped)]

# merge subclasses of celltypes
sample_metadata <- sample_metadata[, celltype.mapped := gsub("[1-9]", "", celltype.mapped)]%>% 
   .[,stage_lineage := paste(stage, celltype.mapped, sep="_")] 
celltypes <- sample_metadata[, unique(celltype.mapped)]
```



# Load data 

# Load global stats per cell

```{r}
acc.stats <- fread(io$acc.stats) %>% .[,c("id_acc","mean")] %>% .[,context:="GC"]
met.stats <- fread(io$met.stats) %>% .[,c("id_met","mean")] %>% .[,context:="CG"]
```

## Load marker peaks from the Multiome 10x data set

```{r}
io$markerPeaks <- "/bi/scratch/Stephen_Clark/tet_chimera_nmtseq/features/filt/marker_peaks.txt.gz"
marker_peaks.dt <- fread(io$markerPeaks) %>% setnames("idx","id") 

# merge subclasses of celltypes
marker_peaks.dt[, celltype := gsub("[1-9]", "", celltype)]
marker_peaks.dt[, unique(celltype)]

# ?? could remove celltypes we don't have from this ??

head(marker_peaks.dt)



```


## Load differential results from the Multiome 10x data set

```{r}
# io$archR.diff.dir <- "/Users/ricard/data/gastrulation_multiome_10x/results/atac/archR/differential/PeakMatrix"
# 
# opts$matrix <- "PeakMatrix"
# opts$min.MeanDiff <- 0.1
# opts$min.FDR <- 1e-1
# opts$celltypes <- c("Gut","Def._endoderm")
# 
# file <- sprintf("%s/%s_%s_vs_%s.txt.gz", io$archR.diff.dir,opts$matrix,opts$celltypes[1],opts$celltypes[2])
# multiome_diff.dt <- fread(file, select = c(1,2,3)) %>% 
#   setnames("idx","id") %>%
#   .[,c("celltypeA","celltypeB"):=list(as.factor(opts$celltypes[1]),as.factor(opts$celltypes[2]))] %>%
#   .[,sig:=FDR<opts$min.FDR & abs(MeanDiff)>=opts$min.MeanDiff] %>%
#   .[,sign:=sprintf("Up in %s",opts$celltypes[1])] %>% 
#   .[MeanDiff<0,sign:=sprintf("Up in %s",opts$celltypes[2])] %>%
#   .[,sign:=as.factor(sign)] %>%
#   setorder(-sig,FDR,na.last = T)
# 
# multiome_diff.dt[,sum(sig),by=c("celltypeA","celltypeB","sign")]
```

## Load feature metadata

```{r}
feature_metadata <- lapply(opts$annos, function(i) 
  fread(
    file = sprintf("%s/%s.bed.gz",io$features.dir,i), 
    select = c("V1"="character", "V2"="integer", "V3"="integer","V5"="character","V6"="factor"))
) %>% rbindlist %>% setnames(c("chr","start","end","id","anno")) 

if (!feature_metadata[1, chr %like% "chr"]) feature_metadata[, chr := paste0("chr", chr)]

feature_metadata[,idx:=sprintf("%s:%s-%s",chr,start,end)]

head(feature_metadata)
```

## Load methylation data

```{r load_metdata, echo=FALSE, include=FALSE}
# met_dt <- lapply(opts$annos, function(n) {
#   fread(
#     sprintf("%s/%s.tsv.gz",io$met_data_parsed,n),
#     showProgress = F, header = F,
#     select = c("V1"="factor","V2"="character","V3"="factor","V4"="integer","V5"="integer","V6"="integer")
#     ) %>% .[V1%in%opts$met.cells] %>% droplevels
# }) %>% rbindlist %>% setnames(c("id_met","id","anno","Nmet","Ntotal","rate"))


met_dt <- paste0(io$met_data_parsed, "/", opts$annos, ".tsv.gz") %>% 
  map(fread) %>% 
  rbindlist()


```

## Load accessibility data

```{r load_accdata, echo=FALSE, include=FALSE}
# acc_dt <- lapply(opts$annos, function(n) {
#   fread(
#     sprintf("%s/%s.tsv.gz",io$acc_data_parsed,n),
#     showProgress = F, header = F,
#     select = c("V1"="factor","V2"="character","V3"="factor","V4"="integer","V5"="integer","V6"="integer")
#     ) %>% .[V1%in%opts$acc.cells] %>% droplevels
# }) %>% rbindlist %>% setnames(c("id_acc","id","anno","Nmet","Ntotal","rate"))

acc_dt <- paste0(io$acc_data_parsed, "/", opts$annos, ".tsv.gz") %>% 
  map(fread) %>% 
  rbindlist()


```

# Fix my stupid errors

```{r}
# met_dt <- met_dt %>% 
#   merge(feature_metadata[,c("id","idx")], by="id") %>% 
#   .[,id:=NULL] %>% setnames("idx","id")
# acc_dt <- acc_dt %>% 
#   merge(feature_metadata[,c("id","idx")], by="id") %>% 
#   .[,id:=NULL] %>% setnames("idx","id")
# feature_metadata %>% 
#   .[,id:=NULL] %>% setnames("idx","id")
# stopifnot(unique(met_dt$id) %in% feature_metadata$id)
# stopifnot(unique(acc_dt$id) %in% feature_metadata$id)
```

# Filter features

```{r}
# met_dt.filt <- met_dt %>% merge(multiome_diff.dt[sig==T,c("id","sign")], by="id")
# acc_dt.filt <- acc_dt %>% merge(multiome_diff.dt[sig==T,c("id","sign")], by="id")
```

```{r}
met_dt.filt <- merge(
  met_dt,
  marker_peaks.dt[,c("id","celltype","score")], 
  by="id"
)

acc_dt.filt <- merge(
  acc_dt, 
  marker_peaks.dt[,c("id","celltype","score")], 
  by="id"
)
```

```{r}
length(unique(met_dt.filt$id))
length(unique(acc_dt.filt$id))
```


# Plot

## Boxplots of (absolute) methylation rates 

```{r}
# to.plot <- met_dt.filt %>%
#   .[,.(rate=100*(sum(Nmet)/sum(Ntotal))),by=c("id_rna","stage_lineage","anno","sign")] %>%
#   .[,stage_lineage:=factor(stage_lineage, levels=opts$stage_lineage)]
# 
# met_dt <- merge(met_dt, sample_metadata[,c("id_met","id_rna","stage_lineage")], by="id_met")

cols <- c(
  "stage_lineage",
  "class",
  "cg_files",
  "gc_files",
  "tdTOM",
  "KDR-Cy7",
  "CD41-BV421"
)

to.plot_met <- merge(
  met_dt.filt,
  sample_metadata[, .(stage_lineage, class, cell = gsub(".tsv.gz", "", basename(cg_files)))],
  by = "cell"
) %>% 
  .[, .(rate = mean(rate)), .(cell, stage_lineage, anno, celltype, class)]


# filter out feature sets that don't match cell types we have
celltypes <- sample_metadata[, unique(celltype.mapped)] %>% .[!is.na(.)]

to.plot_met <- to.plot_met[celltype %in% celltypes]

to.plot_met <- merge(
  to.plot_met, 
  met.stats[, .(cell = id_met, mean)],
  by = "cell"
) %>% 
  .[, scaled := rate / (100*mean)]

##########################################################

# repeat for accessibility

to.plot_acc <- merge(
  acc_dt.filt,
  sample_metadata[, .(stage_lineage, class, cell = gsub(".tsv.gz", "", basename(gc_files)))],
  by = "cell"
) %>% 
  .[, .(rate = mean(rate)), .(cell, stage_lineage, anno, celltype, class)]


# filter out feature sets that don't match cell types we have

to.plot_acc <- to.plot_acc[celltype %in% celltypes]

to.plot_acc <- merge(
  to.plot_acc, 
  acc.stats[, .(cell = id_acc, mean)],
  by = "cell"
) %>% 
  .[, scaled := rate / (100*mean)]





```

```{r}


# features_sub <- c(
#   "Cardiomyocytes",
#   "Erythroid",
#   "Forebrain_Midbrain_Hindbrain" ,
#   "Gut",
#   "ExE_endoderm"
# )

lineages_sub <- to.plot_met[, length(unique(cell)), stage_lineage] %>% 
  .[V1 >= 20, stage_lineage]

features_sub <- gsub("E[1-9].5_", "", lineages_sub)

p <- ggplot(to.plot_met[celltype %in% features_sub & stage_lineage %in% lineages_sub], aes(x = stage_lineage, y = rate, fill = class)) +
  geom_boxplot(outlier.shape=NA, coef=1) +
  geom_point(position = position_jitterdodge(jitter.width = 0.5), size = 0.1, alpha = 0.2) +
  labs(x="", y="Methylation") +
  facet_wrap(~celltype, scales="fixed") +
  #scale_fill_manual(values=opts$colors) +
  # coord_cartesian(ylim=c(5,60)) +
  theme_bw() +
  guides(x = guide_axis(angle = 90)) +
  theme(
    #legend.position = "none",
    axis.text = element_text(color="black")
  )
# pdf(paste0(io$outdir,"/boxplots_acc_top250.pdf"), width=opts$width, height=opts$height/1.5)
print(p)

dir.create(io$outdir, recursive = TRUE)
cowplot::save_plot(paste0(io$outdir, "/met_boxplots.pdf"), p, base_height = 20, base_width = 20)

# dev.off()
```




```{r}


p <- ggplot(to.plot_acc[celltype %in% features_sub & stage_lineage %in% lineages_sub], aes(x = stage_lineage, y = rate, fill = class)) +
  geom_boxplot(outlier.shape=NA, coef=1) +
  geom_point(position = position_jitterdodge(jitter.width = 0.5), size = 0.1, alpha = 0.2) +
  labs(x="", y="Accessibility") +
  facet_wrap(~celltype, scales="fixed") +
  #scale_fill_manual(values=opts$colors) +
  # coord_cartesian(ylim=c(5,60)) +
  theme_bw() +
  guides(x = guide_axis(angle = 90)) +
  theme(
    #legend.position = "none",
    axis.text = element_text(color="black")
  )
# pdf(paste0(io$outdir,"/boxplots_acc_top250.pdf"), width=opts$width, height=opts$height/1.5)
print(p)

dir.create(io$outdir, recursive = TRUE)
cowplot::save_plot(paste0(io$outdir, "/acc_boxplots.pdf"), p, base_height = 20, base_width = 20)

# dev.off()
```

Test
```{r}
# to.plot2 <- to.plot %>% merge(sample_metadata[,c("id_rna","lineage10x")])
# p <- ggplot(to.plot2[stage_lineage=="E7.5_Endoderm"], aes(x=lineage10x, y=rate, fill=lineage10x)) +
#   geom_boxplot(outlier.shape=NA, coef=1) +
#   labs(x="", y="Methylation (%)") +
#   facet_wrap(~sign, nrow=1, scales="fixed") +
#   # coord_cartesian(ylim=c(5,60)) +
#   scale_fill_manual(values=opts$celltype.colors) +
#   theme_bw() +
#   guides(x = guide_axis(angle = 90)) +
#   theme(
#     legend.position = "none",
#     axis.text = element_text(color="black")
#   )
# print(p)
```

# Plot

Define global stats per cell

```{r}
acc.stats <- acc_dt[,.(mean=mean(rate)),by="id_acc"] %>% .[,context:="GC"]
met.stats <- met_dt[,.(mean=mean(rate)),by="id_met"] %>% .[,context:="CG"]
```

```{r}
theme_pub <- function() {
    theme(
      axis.text.x = element_text(size=rel(0.50), color="black"),
      axis.text.y = element_text(size=rel(0.75), color="black"),
      axis.title = element_text(size=rel(0.75), color="black"),
      legend.position = "none"
    )
}
```

```{r}
for (i in unique(marker_peaks.dt$celltype)) {
  
  ## Prepare data
  
  to.plot.met <- met_dt %>% 
    .[id%in%marker_peaks.dt[celltype==i,id]] %>%
    .[,.(rate=100*(sum(Nmet)/sum(Ntotal))),by=c("id_met")] %>%
    merge(sample_metadata[,c("id_met","id_rna","lineage10x","stage_lineage")], by="id_met") %>%
    merge(met.stats,by="id_met") %>% .[,norm_rate:=rate/mean]  %>%
    # .[,lineage10x:=factor(lineage10x, levels=opts$)]
    .[,stage_lineage:=factor(stage_lineage, levels=opts$stage_lineage)]
  
  to.plot.acc <- acc_dt %>% 
    .[id%in%marker_peaks.dt[celltype==i,id]] %>%
    .[,.(rate=100*(sum(Nmet)/sum(Ntotal))),by=c("id_acc")] %>%
    merge(sample_metadata[,c("id_acc","id_rna","lineage10x","stage_lineage")], by="id_acc") %>%
    merge(acc.stats,by="id_acc") %>% .[,norm_rate:=rate/mean]  %>%
    # .[,lineage10x:=factor(lineage10x, levels=opts$)]
    .[,stage_lineage:=factor(stage_lineage, levels=opts$stage_lineage)]
  
  ## Plot
  
  p.met.unscaled <- ggboxplot(to.plot.met, x = "stage_lineage", y = "rate", fill="stage_lineage", coef=0.5, outlier.shape=NA) +
    scale_fill_manual(values=opts$stagelineage.colors) +
    labs(x="", y="Methylation (%)") +
    guides(x = guide_axis(angle = 90)) +
    coord_cartesian(ylim=c(0,100)) +
    theme_pub()
  p.met.scaled <- ggboxplot(to.plot.met, x = "stage_lineage", y = "norm_rate", fill="stage_lineage", coef=0.5, outlier.shape=NA) +
    geom_hline(yintercept=1, linetype="dashed", alpha=0.75, size=0.75) +
    scale_fill_manual(values=opts$stagelineage.colors) +
    labs(x="", y="Methylation (% relative to bgd)") +
    guides(x = guide_axis(angle = 90)) +
    coord_cartesian(ylim=c(0.1,1.75)) +
    theme_pub()
  
  p.acc.unscaled <- ggboxplot(to.plot.acc, x = "stage_lineage", y = "rate", fill="stage_lineage", coef=0.5, outlier.shape=NA) +
    scale_fill_manual(values=opts$stagelineage.colors) +
    labs(x="", y="Accessibility (%)") +
    coord_cartesian(ylim=c(20,70)) +
    guides(x = guide_axis(angle = 90)) +
    theme_pub()
  
  p.acc.scaled <- ggboxplot(to.plot.acc, x = "stage_lineage", y = "norm_rate", fill="stage_lineage", coef=0.5, outlier.shape=NA) +
    geom_hline(yintercept=1, linetype="dashed", alpha=0.75, size=0.75) +
    scale_fill_manual(values=opts$stagelineage.colors) +
    labs(x="", y="Accessibility (% relative to bgd)") +
    coord_cartesian(ylim=c(0.6,1.8)) +
    guides(x = guide_axis(angle = 90)) +
    theme_pub()
  
  p <- cowplot::plot_grid(plotlist=list(p.met.unscaled, p.met.scaled, p.acc.unscaled, p.acc.scaled), nrow = 2, ncol=2)
  pdf(sprintf("%s/%s_markerPeaks_scnmt_metacc.pdf",io$outdir,i), width=11, height=8)
  print(p)
  dev.off()
}
```